Thermodynamic assessment of the Kalina Split Cycle

### 2013-08-08 paper_SPLIT#2.tex
Basic addition

\nomenclature[A]{ASPEN}{Advanced System for Process Engineering}
\nomenclature[A]{WHR}{Waste Heat Recovery}
\nomenclature[A]{ORC}{Organic Rankine Cycle}
\nomenclature[A]{SC}{Split Cycle}
\nomenclature[A]{SUP}{Superheated state}
\nomenclature[A]{SUB}{Subcooled state}
\nomenclature[M]{$\dot{m}$}{Mass flow rate (kg/s)}
\nomenclature[M]{$h$}{Specific enthalpy (kJ/kg)}
\nomenclature[M]{$x$}{Concentration by mass (-)}
\nomenclature[M]{$P$}{Pressure (bar)}
\nomenclature[M]{$q$}{Vapour quality (-)}

%%%%%%%%% SECTION: METHODOLOGY %%%%%%%%%

@@

\subsection{Reference Kalina cycle}
\label{sec:kalina_cycle}
\begin{figure*}[htbp]
\centering
\includegraphics[width=0.9\linewidth]{Drawing_Kalina_Baselinevsd.pdf}
\caption{Sketch of the Kalina reference process with reheat}
\label{fig:kalina_cycle}
\end{figure*}

@@

The gradient of the evaporation temperature curve can to some degree be adjusted to the temperature profile of the heat source by selecting the optimal composition of each of the two streams, as illustrated in Fig. \ref{fig:theory}. The line from (25,31) to the point ($T_{r,b}$)\nomenclature[S]{r}{Rich ammonia concentration} \nomenclature[S]{b}{Bubble point}represents the preheat stage. From here to points (1), (2) and (3) the fully drawn line represents the heat transfer when using the Split-cycle configuration. The upper dashed line represents how the heat transfer would be if the two streams were combined into a single stream. The point ($T_{b}$) signifies the bubble point of the combined stream, and it is clear that the temperature difference at the pinch point is much smaller and possibly violated in this case. The lower dashed line from ($T_{r,b}$) through (2a) to (2), shows how the heat transfer would occur if only the rich stream concentration was used. Evaporation would take place at lower temperatures possibly leading to a lower thermal efficiency.

@@

\section{Methodology}
\label{sec:methods}

@@

	\subsubsection{Energy balance and forms}

@@

	where $\dot{m}$ is the mass flow rate of a stream of matter, $\dot{Q}_{k}$ and $\dot{W}_{cv}$ the time rates of energy transfer by heat and work ($\dot{Q}\ge0$ indicates heat transfer to the system, $\dot{W}\ge0$ work done by the system), $h$ the specific total enthalpy of a stream of matter. The subscripts $in$ and $out$ denote the inlet and outlet of the system, $k$ the boundary of the component of interest and $cv$ the control volume.

	\nomenclature[0]{h}{specific total enthalpy (mass) [J/kg]}
	\nomenclature[S]{k}{component}
	\nomenclature[S]{cv}{control volume}
	\nomenclature[S]{in}{inlet}
	\nomenclature[S]{out}{outlet}

	The specific energy associated with a stream of matter $\hat{h}_k$ is then defined as:

	\begin{equation}
	\hat{h}_j=h_j-h_{j,0}+HHV_j
	\end{equation}

	where $h_j$ and $h_{j,0}$ are the total specific enthalpies of the stream $j$ at a given temperature and pressure, and at reference conditions, respectively, and HHV the higher heating value. Contributions of the potential and kinetic effects are neglected.

	\nomenclature[0]{$\hat{h}$}{specific energy (mass) [J/kg]}
	\nomenclature[A]{HHV}{Higher Heating Value [J/kg]}

	\subsubsection{Application}

	The energy balance for the processing plant of the oil and gas facility can be expressed as:

	\begin{equation}
		\dot{\hat{H}}_{feed}+\dot{Q}_{heat}+\dot{W}_{UT}=\sum_k\dot{\hat{H}}_k+\dot{Q}_{cool}
	\end{equation}

	The left-hand side (LHS) term represents, on a time rate basis, the energy associated with the feed entering the processing plant (i.e. reservoir fluid) $\dot{\hat{H}}_{feed}$ and the energy transfers with the heating medium $\dot{Q}_{heat}$ and power $\dot{W}_{UT}$ from the utility plant that are consumed within the separation and treatment modules. The right-hand side (RHS) term denotes the energy of the outlet streams of the processing plant (i.e. oil, gas, condensate, flared gas, fuel gas, produced water) $\sum_k\dot{\hat{H}}_k$ and the energy discarded with the cooling medium discharged to the sea $\dot{Q}_{cool}$.

	In this work, the reference conditions for the enthalpy calculations are taken to be 8 $^{\circ}$C and an absolute pressure of 1 atmosphere. Higher heating values are determined for the same pressure but for a temperature of 15 $^{\circ}$C -- variations of the calorific values for this temperature difference are neglected.

@@

	\emph{Exergy} may be defined as the `maximum theoretical useful work (shaft work or electrical work) as the system is brought into complete thermodynamic equilibrium with the thermodynamic environment while the system interacts with this environment only' \cite{BejanAdrian;TsatsaronisGeorge;Moran1996,Moran1998,Thermoeconomics2001}. A system in thermal and mechanical equilibrium with the environment is called in \emph{environmental state} and in \emph{dead state} when it is also in chemical equilibrium \cite{Kotas1995}. In this work, the discussions on exergetic efficiencies focus exclusively on the exergy associated with mass and energy transfers, namely the `flow exergy'. Unlike energy, exergy is not conserved in real processes -- some is destroyed due to internal irreversibilities. On a time rate form and for a control volume with in- and outgoing flows, the exergy balance is expressed as \cite{BejanAdrian;TsatsaronisGeorge;Moran1996}:

@@

	where $\dot{E}_d$, $\dot{E}_{in}$ and $\dot{E}_{out}$ are the destroyed, inflowing and outflowing exergy rates, respectively. $e$ is the specific exergy of a stream of matter, $T_0$ and $T_k$ are the environmental and instantaneous temperatures.  The exergy destruction rate can also be calculated from the Gouy-Stodola theorem, which is, on a time rate form, expressed as \cite{Bejan2006}:

	\nomenclature[0]{e}{specific exergy (mass) [J/kg]}
	\nomenclature[0]{$\dot{S}$}{Entropy rate [MW/K]}
	\nomenclature[S]{gen}{generation}
	\nomenclature[0]{p}{pressure [MPa]}
	\nomenclature[0]{$\dot{E}$}{Exergy rate [MW]}
	\nomenclature[S]{d}{destruction} 
	\nomenclature[S]{l}{loss} 
	\nomenclature[S]{0}{dead state} 
	\nomenclature[P]{$^{*}$}{relative}	
	\nomenclature[0]{T}{Temperature [K]}

@@

	Exergy destruction is also called `internal exergy losses', as they are related to the irreversibilities taking place within the system, i.e. \emph{inside} the control volume under study. The term exergy loss, written $\dot{E}_l$ on a time rate basis \cite{BejanAdrian;TsatsaronisGeorge;Moran1996} refers to the exergy discharged to the environment without any practical use (e.g. exergy content of exhaust gases from a gas turbine -- exergy transferred to the cooling water), i.e. \emph{outside} the control volume under study. This waste exergy is destroyed when mixed irreversibly with the environment and is denoted `external exergy losses' in the works of Szargut and Kotas \cite{Szargut1998,Kotas1995}.

@@

	\nomenclature[P]{ph}{physical} 
	\nomenclature[P]{ch}{chemical} 
	\nomenclature[P]{kn}{kinetic} 
	\nomenclature[P]{pt}{potential} 
	\nomenclature[0]{$\bar{e}$}{specific exergy (molar) [J/mol]} 
	\nomenclature[0]{y}{component/sub-system exergy ratio}

@@

	where $h$ and $s$ are the specific enthalpy and entropy of a stream of matter per unit-of-mass, respectively.

@@

	\nomenclature[0]{$h$}{specific enthalpy (mass), J/kg}
	\nomenclature[0]{$s$}{specific entropy (mass), J/(kg$\cdot$K)}
	\nomenclature[S]{j}{stream}

@@

	Kinetic and potential contributions on the flow exergies were assumed to be negligible toward physical and chemical exergies.

	Exergy transported with work is equal to its energy and depends on the system and environmental temperatures when it is transferred with heat:

@@

	\subsubsection{Exergy performance indicators}
	The values of the exergy destruction and loss rates provide information on the system inefficiencies, and other performance parameters related to these variables were developed \cite{BejanAdrian;TsatsaronisGeorge;Moran1996,Kotas1980,Kotas1980a,Kotas1995}. They can illustrate the possibilities for improvement and indicate the components and sub-systems on which attention should be focused:

	\item The exergy destruction ratio $y_{d}^{*}$, defined as the ratio of the exergy destruction rate $\dot{E}_{d,k}$ within a specific process component $k$ to the exergy destruction rate within the whole system $\dot{E}_{d}$:  
	\begin{equation} y_{d}^{*}=\frac{\dot{E}_{d,k}}{\dot{E}_{d}} \end{equation}

	\item The exergy loss ratio $y_{l}^{*}$, defined as the ratio of the exergy loss rate $\dot{E}_{l,k}$ within a specific process component $k$ to the exergy loss rate of the whole system $\dot{E}_{l}$:  
	\begin{equation} y_{l}^{*}=\frac{\dot{E}_{l,k}}{\dot{E}_{l}} \end{equation}

@@

	It should be emphasised that the fuel and product exergies are not necessarily equal to the exergy flows entering $\dot{E}_{in,k}$ and leaving $\dot{E}_{out,k}$ the component/sub-system, but rather the exergy flows that are useful (product) or used (fuel). The definitions of exergetic fuels and products for the components and sub-systems investigated within this study are extensively discussed in Bejan et al. \cite{BejanAdrian;TsatsaronisGeorge;Moran1996} and in Kotas \cite{Kotas1980,Kotas1980a,Kotas1995}.

	\begin{align}
		\delta_{d,k}=&\frac{\dot{E}_{d,k}}{\dot{E}_{f,tot}} \\
		\delta_{l,j}=&\frac{\dot{E}_{l,j}}{\dot{E}_{f,tot}}
	\end{align}

	\item Alternatively, for systems such as physical separation plants or that include dissipative devices, the relationship between the irreversibilities of a system and its total exergy input can be expressed with the exergy loss ratio $\lambda$ \cite{Kotas1980a,Kotas1995}. It represents the proportion of the total exergy flowing into the control volume that is lost through irreversibilities and is denoted irreversibility ratio in the rest of the study to avoid confusion with the exergy loss ratio $y^*_l$ defined in Bejan et al. \cite{BejanAdrian;TsatsaronisGeorge;Moran1996}.

	\begin{equation}
			\lambda_k=\frac{\dot{I}_k}{\dot{E}_{in,k}}
	\end{equation}

	where $\dot{I}$ is the rate of irreversibilities of the system $k$ and is equal to the sum of the exergy destruction and losses for a given system.

@@

	\nomenclature[G]{$\lambda$}{irreversibility ratio}	
	\nomenclature[0]{$\dot{I}$}{Irreversibilities rate [MW]}		
	\nomenclature[G]{$\varepsilon$}{exergy efficiency} 
	\nomenclature[G]{$\delta$}{efficiency defect}	
	\nomenclature[S]{f}{fuel} 
	\nomenclature[S]{p}{product}
	\nomenclature[S]{tot}{total}

	\subsubsection{Application}

	Similarly, the exergy accounting for the Kalina Split Cycle can be expressed as:

@@

	$\dot{E}^{Q}_{heat}$ represents the exergy decrease of the heat source, which is used as fuel to the power cycle system. $\dot{E}^{W}$ denotes the exergy associated with the net power produced, $\dot{E}^{Q}_{cool}$ the exergy discharged to the environment with the exergy gain of the cooling water, and $\dot{E}_{d}$ the exergy destroyed in the plant. The dead state is taken to the environmental conditions and the reference environment of Szargut \cite{Szargut1998} is considered for the chemical exergy calculations.

>>>>

\nomenclature[A]{ASPEN}{Advanced System for Process Engineering}
\nomenclature[A]{WHR}{waste heat recovery}
\nomenclature[A]{ORC}{Organic Rankine Cycle}
\nomenclature[A]{SC}{Split Cycle}
\nomenclature[A]{SUP}{superheated state}
\nomenclature[A]{SUB}{sub-cooled state}

%%%%%%%%% SECTION: SYSTEM DESCRIPTION %%%%%%%%%

@@

\subsection{Reference Kalina cycle}
\label{sec:kalina_cycle}
\begin{figure*}[htbp]
\centering
\includegraphics[width=0.9\linewidth]{Drawing_Kalina_Baseline.pdf}
\caption{Sketch of the Kalina reference process with reheat}
\label{fig:kalina_cycle}
\end{figure*}

@@

The gradient of the evaporation temperature curve can to some degree be adjusted to the temperature profile of the heat source by selecting the optimal composition of each of the two streams, as illustrated in Fig. \ref{fig:theory}. The line from (25,31) to the point ($T_{r,b}$)\nomenclature[S]{r}{rich ammonia concentration} \nomenclature[S]{b}{bubble point}represents the preheat stage. From here to points (1), (2) and (3) the fully drawn line represents the heat transfer when using the Split-cycle configuration. The upper dashed line represents how the heat transfer would be if the two streams were combined into a single stream. The point ($T_{b}$) signifies the bubble point of the combined stream, and it is clear that the temperature difference at the pinch point is much smaller and possibly violated in this case. The lower dashed line from ($T_{r,b}$) through (2a) to (2), shows how the heat transfer would occur if only the rich stream concentration was used. Evaporation would take place at lower temperatures possibly leading to a lower thermal efficiency.

@@

\section{Methods}
\label{sec:methods}

	\subsection{Process modelling, simulation and validation}

		\subsubsection{Modelling}

		\subsubsection{Simulation}

		\subsubsection{Validation}

@@

@@

	where $\dot{Q}_{k}$ and $\dot{W}_{cv}$ the time rates of energy transfer by heat and work ($\dot{Q}\ge0$ indicates heat transfer to the system, $\dot{W}\ge0$ work done by the system). The subscripts $in$ and $out$ denote the inlet and outlet of the system, $k$ the boundary of the component of interest and $cv$ the control volume. $h$ is the specific enthalpy of the stream under consideration, and can be defined as the change in enthalpy observed when it is brought from its temperature and pressure to the reference conditions, and is decomposed into the reference elements in standard state.

	\nomenclature[0]{h}{specific total enthalpy (mass), J/kg}
	\nomenclature[S]{k}{component}
	\nomenclature[S]{cv}{control volume}
	\nomenclature[S]{in}{inlet}
	\nomenclature[S]{out}{outlet}

	The enthalpy change due to temperature and pressure variations is called the specific physical enthalpy $h^{ph}$ and depends on the reference state, fixed in this work to the environmental conditions $T_0$ and $p_0$. The enthalpy change due to chemical deviations with the environment is called the chemical enthalpy $h^{ch}$ and depends on the choice of the reference species and reaction. The reference environment proposed by Szargut is considered in this work. The chemical enthalpy is therefore equal to the enthalpy of devaluation $h_{dev}$ \cite{Kotas1995}, which can be interpreted as the energy released from a given substance interacting with common chemical species present in the environment \cite{Szargut1988}.

	\nomenclature[0]{$h$}{specific enthalpy (mass), J/kg}
	\nomenclature[0]{$h_{dev}$}{specific enthalpy of devaluation (mass), J/kg}

	The energy efficiency, or thermal efficiency $\eta_{th}$, is therefore defined as:

	\begin{equation}
		\eta_{th} = \frac{\dot{W}}{\dot{Q}_{heat}} 
	\end{equation}

	The thermal efficiency is limited by the cold and hot reservoir temperatures $T_{cold}$ and $T_{hot}$, and this upper bound is given by the Carnot efficiency $\eta_{carnot}$:

	\begin{equation}
		\eta_{carnot} = 1-\frac{T_{cold}}{T_{hot}} 
	\end{equation}

@@

	\emph{Exergy} may be defined as the `maximum theoretical useful work (shaft work or electrical work) as the system is brought into complete thermodynamic equilibrium with the thermodynamic environment while the system interacts with this environment only'. Exergy is not conserved in real processes as it is destroyed because of internal irreversibilities. The exergy balance is expressed as \cite{BejanAdrian;TsatsaronisGeorge;Moran1996}:

@@

	where $\dot{E}_d$, $\dot{E}_{in}$ and $\dot{E}_{out}$ are the exergy destruction, inflows and outflows, respectively. $e$ is the specific exergy of a material stream, $T_0$ and $T_k$ are the environmental and instantaneous temperatures. The exergy destruction rate can be calculated from the Gouy-Stodola theorem, which is, on a time rate form, expressed as \cite{Bejan2006}:

	\nomenclature[0]{$e$}{specific exergy (mass), J/kg}
	\nomenclature[0]{$\dot{S}$}{entropy rate, W/K}
	\nomenclature[S]{gen}{generation}
	\nomenclature[0]{$p$}{pressure, Pa}
	\nomenclature[0]{$\dot{E}$}{exergy rate, W}
	\nomenclature[S]{d}{destruction} 
	\nomenclature[S]{l}{loss} 
	\nomenclature[S]{0}{dead state} 
	\nomenclature[P]{$^{*}$}{relative}	
	\nomenclature[0]{$T$}{temperature, K}

@@

	Exergy destruction is also called `internal exergy losses', as they are related to the irreversibilities taking place within the system, i.e. \emph{inside} the control volume under study. This corresponds, for instance, to the inefficiencies of the turbines and pumps. The term exergy loss, also called `external exergy losses' and written $\dot{E}_l$, refers to the exergy released to the environment without any practical use, i.e. \emph{outside} the control volume under study, such as the exergy gain of the cooling water \cite{Szargut1998,Kotas1995,BejanAdrian;TsatsaronisGeorge;Moran1996}.

@@

	\nomenclature[P]{ph}{physical} 
	\nomenclature[P]{ch}{chemical} 
	\nomenclature[P]{kn}{kinetic} 
	\nomenclature[P]{pt}{potential} 
	\nomenclature[0]{$\bar{e}$}{specific exergy (molar), J/mol} 
	\nomenclature[0]{$y$}{component/sub-system exergy ratio}

@@

	where $s$ is the specific entropy of a stream of matter per unit-of-mass, respectively.

@@

	\nomenclature[0]{$s$}{specific entropy (mass), J/(kg$\cdot$K)}
	\nomenclature[S]{j}{stream}

@@

	Kinetic and potential contributions are not considered in this work. Exergy transported with work is equal to its energy and depends on the system and environmental temperatures when it is transferred with heat:

@@

	The tracing of the energy and exergy flows provides information on the system transformations and inefficiencies: several performance parameters were developed to illustrate the possibilities for improvements and illustrate the components on which improvement efforts should focus \cite{BejanAdrian;TsatsaronisGeorge;Moran1996,Kotas1980,Kotas1980a,Kotas1995}.

	\begin{itemize}	
		\item The exergy destruction ratio $y_{d}^{*}$ is defined as the ratio of the exergy destruction rate $\dot{E}_{d,k}$ within a specific component $k$ to the total exergy destruction rate in the overall system $\dot{E}_{d}$:  
	\begin{equation} y_{d}^{*}=\frac{\dot{E}_{d,k}}{\dot{E}_{d}} \end{equation}

		\item Similarly, the exergy loss ratio $y_{l}^{*}$ is defined as the ratio of the exergy loss rate $\dot{E}_{l,k}$ caused by exergy discharge to the environment from a component or sub-system $k$ to the exergy loss rate of the whole system $\dot{E}_{l}$:  
	\begin{equation} y_{l}^{*}=\frac{\dot{E}_{l,k}}{\dot{E}_{l}} \end{equation}

	\begin{align}
		\delta_{j,k} = \frac{\dot{I}_{j,k}}{\dot{E}_{f,tot}}
	\end{align}

		where $\dot{I}$ is the rate of irreversibilities of the component $k$.

@@

	Fuel and product exergies are not always equal to the exergy flows entering $\dot{E}_{in,k}$ and leaving $\dot{E}_{out,k}$ the component/sub-system: they represent the exergy flows that are useful (product) or used (fuel). The definitions of exergetic fuels and products for the components existing in power cycle processes are introduced and discussed in Kotas \cite{Kotas1980,Kotas1980a,Kotas1995} and in Bejan et al. \cite{BejanAdrian;TsatsaronisGeorge;Moran1996}.

@@

	\nomenclature[G]{$\lambda$}{irreversibility ratio}	
	\nomenclature[0]{$\dot{I}$}{irreversibilities rate, W}		
	\nomenclature[G]{$\varepsilon$}{exergy efficiency} 
	\nomenclature[G]{$\delta$}{efficiency defect}	
	\nomenclature[S]{f}{fuel} 
	\nomenclature[S]{p}{product}
	\nomenclature[S]{tot}{total}

	The exergy accounting for the Kalina Split Cycle can therefore be expressed as:

@@

	where $\dot{E}^{Q}_{heat}$ represents the exergy decrease of the heat source, which is used as fuel to the power cycle system. $\dot{E}^{W}$ denotes the exergy associated with the net power produced, $\dot{E}^{Q}_{cool}$ the exergy discharged to the environment with the exergy gain of the cooling water, and $\dot{E}_{d}$ the exergy destroyed in the plant. The dead state is taken to the environmental conditions and the reference environment of Szargut \cite{Szargut1998} is considered for the chemical exergy calculations.
